---
title: 'DATA 622 - Homework #1'
author: "Zach Alexander"
date: "2/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Libraries used


```{r libraries, warning=False}
library(palmerpenguins)
library(dplyr)
library(ggplot2)
```

##### Part 1: Logistic Regression with binary outcome

*The penguin dataset has a 'species' column. Please check how many categories you have in the species column. Conduct whatever data manipulation you need to do to be able to build a logistic regression with binary outcome. Please explain your reasoning behind your decision as you manipulate the outcome/dependent variable (species).*

First, we'll save the penguin dataset as a tibble so we can manipulate it easier in the next steps:

```{r}
penguins <- tibble(penguins)
```

Then, as mentioned above, we'll first have to identify the number of categories that make up the `species` column. We can do this by running the syntax below:

```{r}
penguins %>% count(species)
```

As we can see above, it looks like there are three species present in this dataset. Therefore, in order to create a logistic regression with a binary outcome, we'll have to determine which two species should be combined. We can do this by first taking a look at the 

##### Part 2: Logistic Regression evaluation (AUC, TPR, FPR, TNR, FNR and Accuracy)
